Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2015-11-29 rmarkdown/prelims_paper.Rnw
finished first set of corrections to prelim paper

ChIP-exo is a modification of the ChIP-seq protocol for high
resolution mapping of transcription factor binding sites. Although
many aspect of the ChIP-exo data analysis are similar to those of
ChIP-Seq, ChIP-exo presents a number of unique challenges. We present
a quality control pipeline that analyzes ChIP-exo's strand imbalance
and library complexity. Assessment of these biases and artifacts are
facilitated through diagnostic plots and summary statistical that
compare a large portion of the genome as partitioned into island with
high depth regions.

We systematically evaluated diverse aspects of ChIP-exo and found the
following characteristics: First, ChIP-exo's background is quite
different than ChIP-Seq's. Second, although often assumed in ChIP-exo
data analysis methods, the ``peak pair'' assumption doesn't hold
locally in real ChIP-exo data. Third, we compared Paired End (PE)
ChIP-Seq with ChIP-exo and found that both protocols are comparable in
resolution and sensitivity for closely located binding event, but as
the distance between binding event increases ChIP-exo shows higher
sensitivity than PE ChIP-Seq. Finally, for a given sequencing depth,
ChIP-exo provides higher sensitivity, specificity and spatial
resolution than PE ChIP-Seq.

@@

\[
\mbox{NSC} = \frac{ \max_\delta y(\delta)}{\min_\delta y(\delta)}
\]

where $y(\delta)$ is strand cross-correlation defined as in (\ref{scc}).

@@

\begin{align}
  y(\delta) = \sum_c w_c r\left[ n_c^+ \left(x + \frac{\delta}{2}
    \right), n_c^- \left( x- \frac{\delta}{2} \right)\right]
\label{scc}
\end{align}

@@

Given a peak region with $m$ positions, with $n$ DNA reads and that
the region is generated by a known $g^*$ amount of binding
events. Each read is described by the start position where the 5' end
is aligned and the strand from which the fragment is sequenced. Since
only one end is sequenced, then the fragment length is not
observed. dPeak assumes that the strand $D_i$ follows a Bernoulli
distribution with known parameter $p_D$, the binding event is not
observed and it follows a
$\mbox{Multinomial}(\pi_0,\pi_1,\cdots,\pi_{g^*})$. And the 5' end
$R_i$ for the $i$-th fragment is generated according to the following
procedure:

\begin{itemize}
\item If the read was sequenced from the forward strand ($D_i = 1$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1, m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g - \delta , \sigma^2)$
  \end{itemize}
\item If the read was sequenced from the backward strand ($D_i =
  0$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1 , m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g + \delta , \sigma^2)$
  \end{itemize}

@@

For this work, we considered a new strategy where the $mu_g$'s are
initialized by partitioning the region into $g^*$ bins and finding the
coverage's local maxima in each bin and we initialize $\delta$ and
$\sigma$ by ranking the regions by the signal strength or an empirical
quantity (as the number of ChIP tag counts in the region), fit the
model for the top $N_{\text{top}}$ regions and then use as initial
values the average of the $\sigma$ and $\delta$ estimates for those
regions.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
  \caption{Flowchart of the ChIP-exo Quality control pipeline.}
  % The genome is partitioned into several regions by removing the empty
  % segments and for each regions we calculate a collection of QC
  % indicators, then for each region $R_k$ a vector of summary
  % statistics $T(R_k)$ is calculated.
  \label{fig:qcdiagram}
\end{figure}

@@

\begin{align}
  y(\delta) = r\left[ n^+ \left(x + \frac{\delta}{2}
    \right), n^- \left( x- \frac{\delta}{2} \right)\right]
\nonumber
\end{align}

>>>>

ChIP-exo is a modification of the ChIP-seq protocol for high
resolution mapping of transcription factor binding sites. Although
many aspect of the ChIP-exo data analysis are similar to those of
ChIP-Seq, ChIP-exo presents a number of unique challenges. We present
a quality control pipeline that analyzes ChIP-exo's strand imbalance
and library complexity. Assessment of these biases and artifacts are
facilitated through diagnostic plots and summary statistics calculated
over regions of the genome with varying levels of coverage.

We systematically evaluated diverse aspects of ChIP-exo and found the
following characteristics: First, ChIP-exo's background is quite
different than ChIP-Seq's. Second, although often assumed in ChIP-exo
data analysis methods, the ``peak pair'' assumption does not hold
locally in actual ChIP-exo data. Third, we for the first time compared
Paired End (PE) ChIP-Seq with ChIP-exo and found that both protocols
are comparable in resolution and sensitivity for closely located
binding event, but as the distance between binding event increases
ChIP-exo shows higher sensitivity than PE ChIP-Seq. Finally, at fixed
sequencing depths, ChIP-exo provides higher sensitivity, specificity
and spatial resolution than PE ChIP-Seq.

@@

\[
\mbox{NSC} = \frac{ \max_\delta y(\delta)}{\min_\delta y(\delta)},
\]
where $y(\delta)$ is strand cross-correlation defined as in (\ref{scc}).

@@

\begin{align}
  y(\delta) = \sum_c w_c r\left[ n_c^+ \left(x + \frac{\delta}{2}
    \right), n_c^- \left( x- \frac{\delta}{2} \right)\right],
\label{scc}
\end{align}

@@

Given a peak region with $m$ positions, with $n$ DNA reads and that
the region is generated by a known $g^*$ amount of binding
events. Each read is described by the start position where the 5' end
is aligned and the strand from which the fragment is sequenced. Since
only one end is sequenced, then the fragment length is not
observed. dPeak assumes that the strand $D_i$ follows a Bernoulli
distribution with known parameter $p_D$, the binding event is not
observed and it follows a
$\mbox{Multinomial}(\pi_0,\pi_1,\cdots,\pi_{g^*})$. Then the observed
data, i.e., the 5' end $R_i$ for the $i$-th fragment is generated
according to the following procedure:

\begin{itemize}
\item If the read is sequenced from the forward strand ($D_i = 1$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1, m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g - \delta , \sigma^2)$
  \end{itemize}
\item If the read is sequenced from the backward strand ($D_i =
  0$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1 , m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g + \delta , \sigma^2)$
  \end{itemize}

@@

For this work, we considered a new strategy where the $\mu_g$'s are
initialized by partitioning the region into $g^*$ bins and finding the
coverage's local maxima in each bin we initialize $\delta$ and
$\sigma$ by ranking the regions by the signal strength or an empirical
quantity (as the number of ChIP tag counts in the region), fit the
model for the top $N_{\text{top}}$ regions and then use as initial
values the average of the $\sigma$ and $\delta$ estimates for those
regions.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
  \caption{Flowchart of the ChIP-exo quality control pipeline.}
  % The genome is partitioned into several regions by removing the empty
  % segments and for each regions we calculate a collection of QC
  % indicators, then for each region $R_k$ a vector of summary
  % statistics $T(R_k)$ is calculated.
  \label{fig:qcdiagram}
\end{figure}

@@

\begin{align}
  y(\delta) = r\left[ n^+ \left(x + \frac{\delta}{2}
    \right), n^- \left( x- \frac{\delta}{2} \right)\right],
\nonumber
\end{align}

### 2015-11-29 rmarkdown/slides.Rnw
finished first set of corrections to prelim paper

\subsection{Quality Control pipeline}

\begin{frame}[plain,t]

\begin{figure}[H]
  \centering
  \includegraphics[width = .75\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
\end{figure}

{\small
  \begin{itemize}
  \item $\mbox{ARC} = \frac{\text{Nr. of reads in the
        region}}{\text{Width of the region}}$
  \item $\mbox{URCR} = \frac{\text{Nr. of reads mapped to only one
        position in the region}}{\text{Nr. of reads in the region}}$
  \item $\mbox{local-NSC}$
  \item $\mbox{FSR} = \frac{\text{Nr. of fwd. strand reads in region}}{\text{Nr. of reads in region}}$
  \end{itemize}  
}

@@

\only<3>{
{\small
where:
\begin{itemize}
\item high - regions with nr. of unique positions $ > 100$
\item med - regions with nr. of unique positions in $(50,100)$
\item low - regions with nr. of unique positions in $(20,100)$
\end{itemize}
}}

@@

\section{Comparison with ChIP-Seq using dPeak}
\begin{frame}[t]
  \frametitle{dPeak model for SET case}

@@

@@

\begin{frame}
  \frametitle{References}

>>>>

\subsection{Quality Control Pipeline}

\begin{frame}[t]
  \frametitle{Quality Control Pipeline}

  \begin{enumerate}
\only<1>{
  \item Partition the genome and generate ChIP-exo islands.
\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{../figs/for_paper/coverage_diagram2_part1.pdf}
\end{figure}
}
\only<2>{
\addtocounter{enumi}{1}
\item Calculate a vector of summary statistics for each islands.
\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{../figs/for_paper/coverage_diagram2_part2.pdf}
\end{figure}
}
\only<3>{
\addtocounter{enumi}{2}
\item Visualize all islands together.
\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{../figs/for_paper/coverage_diagram2_part3.pdf}
\end{figure}
}
  \end{enumerate}
\end{frame}

  % \begin{itemize}
  % \item $\mbox{ARC} = \frac{\text{Nr. of reads in the
  %       region}}{\text{Width of the region}}$
  % \item $\mbox{URCR} = \frac{\text{Nr. of reads mapped to only one
  %       position in the region}}{\text{Nr. of reads in the region}}$
  % \item $\mbox{local-NSC}$
  % \item $\mbox{FSR} = \frac{\text{Nr. of fwd. strand reads in region}}{\text{Nr. of reads in region}}$
  % \end{itemize}

@@

\only<3>{
{\small
where:
\begin{itemize}
\item high - regions with nr. of unique positions $ > 100$
\item med - regions with nr. of unique positions in $(50,100)$
\item low - regions with nr. of unique positions in $(20,50)$
\end{itemize}
}}

@@

\begin{frame}[t]
  \frametitle{dPeak model for SET case}

@@

\section{Comparison with Other Algorithms}
\begin{frame}
\frametitle{Comparison with Other Algorithms}

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/algorithm_resolution.pdf}
\end{figure}

@@

\begin{frame}[allowframebreaks]
  \frametitle{References}
